---
title: "Reproducing and evaluating meta-analyses"
format: 
  html: 
    embed-resources: true
editor: visual    
---

## Pelled et al. (2025): Labor Leads to Love, Right? A Meta-Analysis of the IKEA Effect

Access: <https://linktr.ee/AdvancedStatistics>

1.  Make some reproducibility notes (individually or in small groups)
2.  Attempt to reproduce the analysis (together)
3.  Make some notes about the good/bad things about the paper (individually or in small groups)
4.  Implement an improvement (together)

## Reproducibility notes

**What is/are the primary and secondary outcomes of the meta-analysis?**

**What type of model and effect size did the authors fit to estimate the pooled effect(s)?**

**What was the exact result of overall meta-analysis based on the primary outcome? Report it below.**

**How much heterogeneity was there in the model for the primary outcome? Note down exact values.**

**Which moderators did the authors explore?**

**Choose one moderator and report the results below.**

## Reproducibility attempt

```{r message=FALSE}
# packages: 
library(tidyverse)
library(metafor)
library(emmeans)
```

```{r message = FALSE}
# data: 
ikea_tib <- here::here("data/ikea_data.csv") |> readr::read_csv()
```

```{r}
# data wrangling: 
val_tib <- ikea_tib |> 
  dplyr::filter(Outcome == "Valuation")
```

```{r}
# fit the model (note about alternative estimators)
val_rma <- metafor::rma(yi = `Std diff in means`, sei = `Std Err...7`, data = val_tib)
```

```{r }
# visualise
metafor::forest(val_rma, slab = val_tib$`Study name`)
```

```{r}
# moderation (3d-Printed)
val_rma_mod <- metafor::rma(
  yi = `Std diff in means`, sei = `Std Err...7`, 
  mods = ~ `3d-Printed`,
  data = val_tib)

# emmeans: 
val_emm <- val_rma_mod |> 
  metafor::emmprep() |> 
  emmeans::emmeans(specs = "1", by = "3d-Printed")
```

## Evaluation notes

**What are some things the authors did well?**

**What are some things that could be improved?**

## Improving the analysis

```{r}
# Option 1: 
val_rma_g <- metafor::rma(yi = `Hedges's g`, sei = `Std Err...9`, data = val_tib)


# Option 2: 
val_tib <- val_tib |> 
  dplyr::mutate(vi = `Std Err...7`^2)

val_rmamv <- metafor::rma.mv(
  yi = `Std diff in means`,
  V = vi, 
  random = ~ 1|cluster/`Subgroup within study`, 
  data = val_tib
)

val_rob <- metafor::robust(val_rmamv, cluster = val_tib$cluster)

# Option 3: 
moderate_pb <- metafor::selmodel(
  val_rma, 
  type = "stepfun",
  steps = c(.05, 0.1, 0.5), 
  delta = c(1, 0.7, 0.6, 0.5),
) |> 
  broom::tidy(conf.int = TRUE)

severe_pb <- metafor::selmodel(
  val_rma, 
  type = "stepfun", 
  steps = c(.05, 0.1, 0.5), 
  delta = c(1, 0.5, 0.4, 0.1), 
) |> 
  broom::tidy(conf.int = TRUE) 
```
